---
title: >
  ManyBabies Habituation Project
subtitle: >
  This project aims to explore infant habituation through looking times   using a large harmonised dataset from ManyBabies 1, ManyBabies 3, and ManyBabies 4. Study-relevant materials can be found on the Open Science Framework: https://osf.io/8xm4r
author:
  
- name: Jasmin Dönicke
  affiliation: 
  - University Medical Center Mainz
  email: jasminabuaishah@gmail.com
  orcid: 0009-0006-0721-0790
      
date: 2026-03-21
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
    mainfont: "Atkinson Hyperlegible"
    # mainfont: "IBM Plex Sans"
    link-external-newwindow: true
    # code-link: true ## does seem to interfere with the title... pity.
    # code-tools: true
    theme: 
    - cosmo
    - ummz_theme.scss
execute: 
  cache: false
  echo: true
  eval: true
  output: true
  message: false
  warning: false
editor_options: 
  chunk_output_type: console
  execute:
  error: true
---

# Shape of habituation curve
Explore which functional forms best capture habituation trajectories.


##  Library
```{r}
#| results: hide
# libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
library(splines)
library(broom.mixed)
library(flextable)
library(purrr)
library(mgcv)
```

## Prepare Data

```{r}

#| results: hide
# set environment
setwd("~/Documents/GitHub/ManyBabies-Habituation")

# read CSV
data <- read.csv("final_merged_manybabies.csv")

# assign factors
data <- data %>%
  mutate(across(where(is.character), as.factor),
    age = cut(
      age_days,
      breaks = c(0, 90, 180, 270, 365, 455, 545, 635, 653),  # boundaries in days
      labels = c(
        "< 3 months",
        "3–6 months",
        "6–9 months",
        "9–12 months",
        "12–15 months",
        "15–18 months",
        "18–21 months",
        "> 21 months"),
      right = FALSE),
    age = dplyr::coalesce(age_group_new, age),
    age = factor(age, levels = c(
      "< 3 months",
      "3–6 months",
      "6–9 months",
      "9–12 months",
      "12–15 months",
      "15–18 months",
      "18–21 months",
      "> 21 months")),
    testing_trial = ifelse(trial_number > 0, trial_number, NA), # create variable with only testing trials (Training trials [-2, -1, 0] are excluded because they serve familiarization purposes and are not relevant for the main outcome) 
    log_looking_time = log1p(looking_time), # given the zero inflacted looking times, log1p allows them to be included in the model as log(0) is impossible
    testing_trial_sc = scale(testing_trial), # scale testing_trial to help model estimation
    age_days_sc = as.numeric(scale(age_days))) # scale age_days to help model estimation

trial_mean <- attr(data$testing_trial_sc, "scaled:center")
trial_sd   <- attr(data$testing_trial_sc, "scaled:scale")
data$testing_trial_sc <- as.numeric(data$testing_trial_sc)

# remove NA for complete-case analysis
data_model <- data %>%
  filter(!is.na(looking_time))

```

# Data validation

### Missingness on dependent variable
```{r validation}

# Distribution of zero looking times and NAs
table(is.na(data$looking_time))
table(data$looking_time == 0)

# Are zeros/NAs concentrated in later trials?
aggregate(looking_time == 0 ~ trial_number, data = data, FUN = mean)
agg <- aggregate(is.na(looking_time) ~ trial_number, data = data, FUN = mean)
agg

names(agg)[2] <- "na_prop"

ggplot(agg %>% subset(trial_number >= 1), aes(x = trial_number, y = na_prop)) +
  geom_line() +
  geom_point() +
  labs(x = "Trial number",
       y = "Proportion NA in looking_time")
```

## Simplify Data
We have roughly 55,000 observations. When testing new code, I can use this smaller sample so things run faster (= use data_sample). Once your code is ready, I will apply it to the full dataset (= use data).

```{r}
set.seed(123)

data_sample <- data %>%
  dplyr::sample_n(5000)
```

## Plot Overall Habituation Curve
The plot shows a steel early habituation (steep drop), which flattens at circa trial No. 4. This looks like a possible non-linear relationship!

```{r}

ggplot(data, aes(x = testing_trial, y = looking_time)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)) +
  labs(x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials") +
  theme_minimal()

```

## Plot Individual Habituation Curves

The individual plots show substantial variability in the habituation curves across individuals and time

```{r plot looking times, echo=FALSE}

set.seed(1)
ids <- sample(unique(data$unique_subject_identifier), 200) # to be able to observe trajectories for a subset of participants

df_sub <- subset(data, unique_subject_identifier %in% ids)
df_sub <- df_sub[order(df_sub$unique_subject_identifier, df_sub$testing_trial), ]

ggplot(df_sub,
       aes(x = testing_trial,
           y = looking_time,
           group = unique_subject_identifier,
           color = unique_subject_identifier)) +
  geom_line(alpha = 0.6) +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(c(0, 25))

```

## Missingness

- Before imputing the variables we will need to do a lot of data cleaning for primary_language, race, participant_place_of_birth, parent1_education, parent2_education

### Overall missingness

High degree of missingness for project specific variables like household_size and participant_place_of_birth amongst others as these were not included in some of the projects (e.g., household_size is not included in MB4)

```{r overall missingness, echo=FALSE}

# Overall missingness
missing_summary <- data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Missing_N") %>%
  mutate(Missing_Percent = 100 * Missing_N / nrow(data)) %>%
  flextable()

missing_summary

```

### Missingness by project

```{r}

# for inspection
missing_by_project <- data %>%
  group_by(project_id) %>%
  summarise(across(
    everything(),
    ~ mean(is.na(.)))) %>%
  pivot_longer(
    cols = -project_id,
    names_to = "Variable",
    values_to = "Missing_rate") %>%
  flextable()

missing_by_project
```

## Descriptive statistics and missingness per variables for inspection

```{r desc and missingness, echo=FALSE}

cat_vars <- names(data)[!sapply(data, is.numeric)]
cat_vars <- cat_vars[!cat_vars %in% c("unique_subject_identifier", "stimulus_id", "project_id", "lab_id")]
cont_vars <- names(data)[ sapply(data, is.numeric)]

# categorical variables
cat_summary <- map_dfr(cat_vars, function(var) {

  x <- data[[var]]

  tab <- table(x, useNA = "no")

  bind_rows(tibble(
      Variable = var,
      Level = names(tab),
      N = as.integer(tab),
      Percent = 100 * as.integer(tab) / sum(tab)),
    tibble(
      Variable = var,
      Level = "Missing",
      N = sum(is.na(x)),
      Percent = NA_real_))}) %>%
  flextable()

# continuous variables
cont_summary <- map_dfr(cont_vars, function(var) {

  x <- data[[var]]

  tibble(
    Variable = var,
    Mean = mean(x, na.rm = TRUE),
    SD = sd(x, na.rm = TRUE),
    Missing = sum(is.na(x)))}) %>%
  flextable()

cat_summary
cont_summary

```

## Create Overall Habituation Curve Models
To explore which functional forms best capture habituation trajectories, I model looking time as a function of trial number using a series of regression models. Specifically, I compare linear, quadratic, cubic and other shapes (using splines) relationships the variables.

First, I estimate a linear model to assess whether looking time decreases across trials. I then extend this model by including a quadratic term to capture potential curvature in the habituation trajectory. In addition, I fit flexible non-linear models using spline functions to allow for more complex shapes. Importantly, I always use mixed-models with infant-level random intercepts, to account for repeated measurements within infants.

### check if all participants are nested in only one study

```{r lmer nestedness ID, echo=FALSE}
data %>%
  group_by(unique_subject_identifier) %>%
  summarise(n_projects = n_distinct(project_id)) %>%
  filter(n_projects > 1)
```

### check if labs participated in more than one study

```{r lmer nestedness lab_id, echo=FALSE}

data %>%
  group_by(lab_id) %>%
  summarise(n_projects = n_distinct(project_id)) %>%
  filter(n_projects > 1)

```

### linear model

```{r linear model}
# linear models

model_linear_mixed <- lmer(log_looking_time ~ testing_trial + (1 | project_id) + (1 | lab_id) + (1 | unique_subject_identifier), data = data_model)

model_linear_mixed2 <- lmer(log_looking_time ~ testing_trial + (1 | project_id) + (1 | lab_id) + (1 + testing_trial_sc | unique_subject_identifier), data = data_model)

anova(model_linear_mixed, model_linear_mixed2) # Best fitting model has random slope per ID for testing trial
```

### quadratic model

```{r quadractic model}
# quadratic model
model_quadratic_mixed <- lmer(
  log_looking_time ~ testing_trial + I(testing_trial^2) + (1 | project_id) + (1 | lab_id) + (1 | unique_subject_identifier),
  data = data_model)

model_quadratic_mixed2 <- lmer(
  log_looking_time ~ testing_trial + I(testing_trial^2) + (1 | project_id) + (1 | lab_id) + (1 + testing_trial_sc | unique_subject_identifier),
  data = data_model)

anova(model_quadratic_mixed, model_quadratic_mixed2) # Best fitting model has random slope per ID for testing trial
```

### spline model

#### check number of knots 

```{r number knots, echo=FALSE}

fit_models <- function(df_values, data) {
  models_m1 <- list()
  models_m2 <- list()
  warnings_log <- list()
  
  for (d in df_values) {
    
    # Model 1: intercepts only
    fml_m1 <- as.formula(paste0(
      "log_looking_time ~ ns(testing_trial_sc, df = ", d, ") + ",
      "(1 | project_id) + (1 | lab_id) + (1 | unique_subject_identifier)"))
    
    # Model 2: linear slope for subjects
    fml_m2 <- as.formula(paste0(
      "log_looking_time ~ ns(testing_trial_sc, df = ", d, ") + ",
      "(1 | project_id) + (1 | lab_id) + (1 + testing_trial_sc | unique_subject_identifier)"
    ))
    
    w1 <- character(0)
    w2 <- character(0)
    
    models_m1[[paste0("df_", d)]] <- withCallingHandlers(
      lmer(fml_m1, data = data, control = lmerControl(optimizer = "bobyqa")),
      warning = function(e) { w1 <<- c(w1, conditionMessage(e)); invokeRestart("muffleWarning") }
    )
    
    models_m2[[paste0("df_", d)]] <- withCallingHandlers(
      lmer(fml_m2, data = data, control = lmerControl(optimizer = "bobyqa")),
      warning = function(e) { w2 <<- c(w2, conditionMessage(e)); invokeRestart("muffleWarning") }
    )
    
    warnings_log[[paste0("df_", d)]] <- list(m1 = w1, m2 = w2)
  }
  
  return(list(m1 = models_m1, m2 = models_m2, warnings = warnings_log))
}

result <- fit_models(1:6, data)

# Check warnings per model type
lapply(result$warnings, function(w) list(m1 = w$m1, m2 = w$m2))

# Compare m1 vs m2 for each df
comparisons <- lapply(names(result$m1), function(nm) {
  anova(result$m1[[nm]], result$m2[[nm]])
})
names(comparisons) <- names(result$m1)
comparisons

# ── Model comparison

# Within m1: compare across df values
do.call(anova, unname(result$m1))

# Within m2: compare across df values
do.call(anova, unname(result$m2))

# M1 vs M2 within each df
comparisons <- lapply(names(result$m1), function(nm) {
  cat("\n---", nm, "---\n")
  print(anova(result$m1[[nm]], result$m2[[nm]]))
})

# plot AIC and BIC

aic_bic_df <- data.frame(
  df = 1:6,
  M1_AIC = sapply(result$m1, AIC),
  M2_AIC = sapply(result$m2, AIC),
  M1_BIC = sapply(result$m1, BIC),
  M2_BIC = sapply(result$m2, BIC)) %>%
  pivot_longer(-df, names_to = c("model", "metric"), names_sep = "_") %>%
  rename(value = value)

ggplot(aic_bic_df, aes(x = df, y = value, color = model, linetype = metric, group = interaction(model, metric))) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  scale_x_continuous(breaks = 1:6) +
  theme_minimal() +
  labs(x = "Spline df", y = "Value", color = "Random effects", linetype = "Criterion",
       title = "AIC and BIC across models")
```

#### Check visual fit

```{r model plots, echo=FALSE}

good_models_m1 <- names(result$m1)[sapply(result$m1, function(m) {
  !isSingular(m) && is.null(m@optinfo$conv$lme4$messages)
})]

good_models_m2 <- names(result$m2)[sapply(result$m2, function(m) {
  !isSingular(m) && is.null(m@optinfo$conv$lme4$messages)
})]

newdat <- data.frame(
  testing_trial_sc = seq(
    min(data$testing_trial_sc, na.rm = TRUE),
    max(data$testing_trial_sc, na.rm = TRUE),
    length.out = 100
  )
) %>%
  mutate(testing_trial = testing_trial_sc * trial_sd + trial_mean)

# Verify before plotting
range(newdat$testing_trial)

preds_m1 <- lapply(good_models_m1, function(nm) {
  mod <- result$m1[[nm]]
  nd <- newdat
  nd$fit <- predict(mod, newdata = nd, re.form = NA)
  nd$model <- nm
  nd
})
names(preds_m1) <- good_models_m1

preds_m2 <- lapply(good_models_m2, function(nm) {
  mod <- result$m2[[nm]]
  nd <- newdat
  nd$fit <- predict(mod, newdata = nd, re.form = NA)
  nd$model <- nm
  nd
})
names(preds_m2) <- good_models_m2

plot_df_m1 <- do.call(rbind, preds_m1)
plot_df_m2 <- do.call(rbind, preds_m2)

summary_df <- data %>%
  group_by(testing_trial) %>%
  summarise(
    mean_looking = mean(log_looking_time, na.rm = TRUE),
    sd = sd(log_looking_time, na.rm = TRUE),
    n = n(),
    se = sd / sqrt(n),
    ci = 1.96 * se,
    .groups = "drop"
  )

# ── Plot m1 

ggplot() +
  geom_ribbon(data = summary_df, aes(x = testing_trial, ymin = mean_looking - ci, ymax = mean_looking + ci), alpha = 0.2) +
  geom_point(data = summary_df, aes(x = testing_trial, y = mean_looking), size = 1.5) +
  geom_line(data = summary_df, aes(x = testing_trial, y = mean_looking), linewidth = 0.8) +
  geom_line(data = plot_df_m1, aes(x = testing_trial, y = fit, color = model), linewidth = 1) +
  ylim(1.3, 2.2) +
  theme_minimal() +
  labs(x = "Testing trial", y = "Log looking time", color = "Spline df",
       title = "M1: Intercepts only")

# ── Plot m2

ggplot() +
  geom_ribbon(data = summary_df, aes(x = testing_trial, ymin = mean_looking - ci, ymax = mean_looking + ci), alpha = 0.2) +
  geom_point(data = summary_df, aes(x = testing_trial, y = mean_looking), size = 1.5) +
  geom_line(data = summary_df, aes(x = testing_trial, y = mean_looking), linewidth = 0.8) +
  geom_line(data = plot_df_m2, aes(x = testing_trial, y = fit, color = model), linewidth = 1) +
  ylim(1.3, 2.2) +
  theme_minimal() +
  labs(x = "Testing trial", y = "Log looking time", color = "Spline df",
       title = "M2: Linear slope for subjects")

```

```{r spline model}
model_spline_mixed <- lmer(
  looking_time ~ ns(testing_trial, df = 4) + (1 | project_id) + (1 | lab_id) + (1 | unique_subject_identifier),
  data = data_model)

model_spline_mixed2 <- lmer(
  looking_time ~ ns(testing_trial, df = 4) + (1 | project_id) + (1 | lab_id) + (1 + testing_trial_sc | unique_subject_identifier),
  data = data_model)

anova(model_spline_mixed, model_spline_mixed2) # Best fitting model has random slope per ID for testing trial (model_spline_mixed2)
```

### generalised additive model

```{r GAM}

# GAM 

gam1 <- bam(log_looking_time ~ s(testing_trial_sc, k = 5) +
            s(unique_subject_identifier, bs="re") + s(lab_id, bs = "re") + s(project_id, bs = "re"), # random intercepts
            data=data, method = "fREML") 

#gam2 <- bam(log_looking_time ~ s(testing_trial_sc, k = 5) +
#                     s(unique_subject_identifier, bs="re") +
#                     s(unique_subject_identifier, testing_trial_sc, bs="fs", k = 2) +
#                     s(lab_id, bs="re") +
#                     s(project_id, bs="re"),
#                     data = data,
#                     method = "fREML")

```


## Results

### Linear Model

```{r}

summary(model_linear_mixed2)
lm <- broom.mixed::tidy(model_linear_mixed2, conf.int = TRUE)
lm

lm_slope <- lm$estimate[2]
lm_t_val <- lm$statistic[2]

# back transform baseline
b0 <- fixef(model_linear_mixed2)["(Intercept)"]
baseline <- (exp(b0) - 1)
```

- Looking-time decreases significantly across trials (Slope = `r lm_slope`, t = `r lm_t_val`, which tells us: habituation is happening. 
- There is substantial variability in baseline looking- times across infants, supporting the inclusion of random intercepts.
- Back-transformed predictions indicate that the expected looking time at the first trial was `r baseline` seconds.

### Quadratic Model

```{r}

summary(model_quadratic_mixed2)
qm <- broom.mixed::tidy(model_quadratic_mixed2, conf.int = TRUE)
qm

qm_s_estimate <- lm$estimate[2]
qm_s_t_val <- lm$statistic[2]

qm_qestimate <- lm$estimate[2]
qm_q_t_val <- lm$statistic[2]

```

- There is a significant decrease in looking time across trials (Slope = `r qm_s_estimate`, t = `r qm_s_t_val`) and a significant positive quadratic effect (Slope = `r qm_qestimate`, t = `r qm_q_t_val`).
- This indicates that looking-time decreased rapidly at early trials and then leveled off over time, consistent with a typical habituation trajectory.

### Natural spline Model

```{r}
# A mixed-effects model using natural splines revealed. 
summary(model_spline_mixed2)
broom.mixed::tidy(model_spline_mixed2, conf.int = TRUE)
```

- There is a significant nonlinear decrease in looking time across trials. 
- Splines cannot be interpreted as easily.
- The spline model provided the best fit to the data compared to linear and quadratic models, indicating that habituation follows a nonlinear trajectory characterized by a rapid initial decrease and subsequent leveling off. This pattern is consistent with typical habituation processes.

```{r}

# Plot Spline Model
ggplot(data_model, aes(x = testing_trial, y = looking_time)) +
  geom_smooth(method = "lm", 
              formula = y ~ splines::ns(x, df = 4), 
              se = TRUE) +
  theme_minimal()

```

```{r}

# Plot Spline Model on top of observed data
ggplot(data, aes(x = testing_trial, y = looking_time)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials") +
  geom_smooth(method = "lm", 
              formula = y ~ splines::ns(x, df = 4), 
              se = TRUE) +
  theme_minimal()

# Here, I am showing the average observed habituation pattern and overlaying a smooth model-based curve to see how well a nonlinear trajectory captures it. 
```

### Generalised additive model
- There is a significant nonlinear decrease in looking time across trials. 
- Splines cannot be interpreted as easily.
- The spline model provided the best fit to the data compared to linear and quadratic models, indicating that habituation follows a nonlinear trajectory characterized by a rapid initial decrease and subsequent leveling off. This pattern is consistent with typical habituation processes.

```{r}

# Generalised additive model
summary(gam1)
broom.mixed::tidy(gam1, conf.int = TRUE)

```

## Compare  Overall Habituation Curve Model Fit 

Here, I compare models based on goodness-of-fit criteria (e.g. AIC).

```{r}

AIC(model_linear_mixed2, model_quadratic_mixed2, model_spline_mixed2, gam1)
BIC(model_linear_mixed2, model_quadratic_mixed2, model_spline_mixed2, gam1)

```

Here, I compare models by visual inspection of fitted trajectories. This allows me to evaluate whether habituation follows a linear decline, shows rapid early decreases with later stabilization, or exhibits more complex patterns.

```{r}

data_summary <- data_model %>%
  group_by(testing_trial) %>%
  summarise(
    mean_lt = mean(looking_time, na.rm = TRUE),
    sd_lt   = sd(looking_time, na.rm = TRUE),
    n       = sum(!is.na(looking_time)),
    se_lt   = sd_lt / sqrt(n),
    ci_low  = mean_lt - 1.96 * se_lt,
    ci_high = mean_lt + 1.96 * se_lt,
    .groups = "drop")

# linear model
p1 <- ggplot(data_summary, aes(testing_trial, mean_lt)) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), alpha = 0.2) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Linear Habituation Trajectory",
    x = "Trial Number",
    y = "Mean Looking Time") +
  theme_minimal()

# quadractic model
p2 <- ggplot(data_summary, aes(testing_trial, mean_lt)) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), alpha = 0.2) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2), se = FALSE) +
  labs(
    title = "Quadratic Habituation Trajectory",
    x = "Trial Number",
    y = "Mean Looking Time") +
  theme_minimal()

# spline model
p3 <- ggplot(data_summary, aes(testing_trial, mean_lt)) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), alpha = 0.2) +
  geom_point(size = 2) +
  geom_smooth(method = "lm",
              formula = y ~ splines::ns(x, df = 3),
              se = FALSE) +
  labs(
    title = "Spline Habituation Trajectory",
    x = "Trial Number",
    y = "Mean Looking Time") +
  theme_minimal()

# -----------------------
# 4. GAM model
# -----------------------
p4 <- ggplot(data_model, aes(x = testing_trial, y = looking_time)) +
  geom_point(alpha = 0.2) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, k = 5, bs = "cs"),
    se = TRUE) +
  labs(
    title = "GAM Habituation Trajectory (raw data)",
    x = "Trial Number",
    y = "Looking Time") +
  theme_minimal()

# -----------------------
# Print plots
# -----------------------
p1
p2
p3
p4

```

All models in one plot

```{r}

ggplot(data_summary, aes(x = testing_trial, y = mean_lt)) +
  geom_point(size = 2.2, color = "black") +
  geom_smooth(
    aes(color = "Linear", linetype = "Linear"),
    method = "lm", formula = y ~ x,
    se = FALSE, linewidth = 1) +
  geom_smooth(
    aes(color = "Quadratic", linetype = "Quadratic"),
    method = "lm", formula = y ~ x + I(x^2),
    se = FALSE, linewidth = 1) +
  geom_smooth(
    aes(color = "Spline", linetype = "Spline"),
    method = "lm", formula = y ~ splines::ns(x, df = 3),
    se = FALSE, linewidth = 1) +

  # GAM plot fitted using data from pred_grid
  geom_ribbon(
    data = pred_grid,
    aes(x = testing_trial, ymin = lwr, ymax = upr),
    fill  = "purple",
    alpha = 0.15,
    inherit.aes = FALSE) +
  geom_line(
    data = pred_grid,
    aes(x = testing_trial, y = fit, color = "GAM", linetype = "GAM"),
    linewidth = 1) +
  scale_color_manual(name = "Model",
    values = c("Linear" = "darkgreen",
               "Quadratic" = "blue",
               "Spline" = "red",
               "GAM" = "purple")) +
  scale_linetype_manual(
    name = "Model",
    values = c("Linear" = "dashed",
               "Quadratic" = "solid",
               "Spline" = "dotted",
               "GAM" = "longdash")) +
  scale_x_continuous(breaks = 1:16) +
  labs(x = "Trial Number",
       y = "Mean Looking Time",
       title = "Habituation Trajectories by Model") +
  coord_cartesian(ylim = c(5, 10)) +
  theme_minimal()

```

# Individual differences
Identify factors that explain variability across infants.

- 2.1 What factors moderate differences?
- 2.2 Which moderators are most important?
- 2.3 Do moderators relate to different aspects?
- 2.4 Are there identifiable subgroups?

## Habituation by Age
I will use the age variable that I previously created by harmonizing the two variables age_group_new and age_days
```{r}

ggplot(data %>% dplyr::filter(testing_trial > 0), 
       aes(x = testing_trial, y = looking_time, color = age)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Age"
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```


The plot looks strange: Lets check, how many observations are in each age group in the testing trials!
```{r}
data %>%
  dplyr::filter(testing_trial > 0) %>%
  dplyr::count(age)
```

There are few observations in 15-18 and 18-21 months. I will remove NA category for plotting, and add 15-21 months into a single category to make the sample bigger.
```{r}
data <- data %>%
  mutate(new_age = recode(age,
                              "< 3 months" = "< 3 months",
                              "3–6 months" = "3–6 months",
                              "6–9 months" = "6–9 months",
                              "9–12 months" = "9–12 months",
                              "12–15 months" = "12–15 months",
                              "15–18 months" = "15–21 months",
                              "18–21 months" = "15–21 months",
                              "21+ months" = "21+ months"))

# check new age-variable
data %>%
  dplyr::filter(testing_trial > 0) %>%
  dplyr::count(new_age)
```

Plot again with new variable and without NA group (16 observations), with 95% confidence intervals 
```{r}
ggplot(data %>% dplyr::filter(testing_trial > 0, !is.na(age) ), 
       aes(x = testing_trial, y = looking_time, color = new_age)) +
  geom_smooth(se = TRUE, level = 0.95) +
  scale_color_discrete(
    name = "Age"
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

## Habituation by Sex of Participant
There seems to be no sex difference overall! Maybe if we stratify the data by other factors later on.
```{r}
ggplot(data %>% dplyr::filter(testing_trial > 0, !is.na(participant_sex) ), 
       aes(x = testing_trial, y = looking_time, color = participant_sex)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Sex"
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

## Habituation by Gender of Primary Caregiver
```{r}
# check amount of data per category
data %>%
dplyr::filter(testing_trial > 0) %>%
dplyr::count(primarycaregiver_gender)
# "other" and "OTHER" are just 75 observations, compared to 44000 F and 4500 M, remove "other" due to severely missing data for plotting

ggplot(data %>% dplyr::filter(
         !is.na(primarycaregiver_gender),   
         !primarycaregiver_gender %in% c("other", "OTHER")), 
       aes(x = testing_trial, y = looking_time, color = primarycaregiver_gender)) +
  geom_smooth(se = TRUE, level = 0.95) +
  scale_color_discrete(
    name = "Primary Caregiver Gender"
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

## Habituation by Language Group
```{r}
# check amount of data per category
data %>%
dplyr::filter(testing_trial > 0) %>%
dplyr::count(language_group)
# "other" and "OTHER" are just 71 observations, remove them for plotting

ggplot(data %>% dplyr::filter(!is.na(language_group),
                              !language_group %in% c("other")), 
       aes(x = testing_trial, y = looking_time, color = language_group)) +
          geom_smooth(se = TRUE, level = 0.95) +
          scale_color_discrete(
            name = "Lanuage Group"
          ) +
          labs(
            x = "Trial Number",
            y = "Looking Time (in seconds)",
            title = "Looking Time Across Trials"
          ) +
          scale_x_continuous(
            breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
          ) +
          theme_minimal()
```



# Methodological effects
Assess how methodological choices influence results.

- 3.1 Do differences contribute to variability?
- 3.2 Do they affect interpretation?
- 3.3 Do they affect reliability?


## Habituation by Trial Type
```{r}
# check amount of data per category
data %>%
dplyr::filter(testing_trial > 0) %>%
dplyr::count(trial_type)


ggplot(data, aes(x = testing_trial, y = looking_time, color = trial_type)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Trial Type",
    labels = c(
      "ADS" = "Adult speech",
      "IDS" = "Infant speech",
      "consistent" = "Consistent",
      "inconsistent" = "Inconsistent",
      "nonsocial" = "Non-Social",
      "social" = "Social"
    )
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

## Habituation by Project ID
```{r}
ggplot(data, aes(x = testing_trial, y = looking_time, color = project_id)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Project ID") +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

#### Habituation by Method
```{r}
ggplot(data, aes(x = testing_trial, y = looking_time, color = method)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Method",
    labels = c(
      "eyetracking" = "Eyetracking",
      "hpp" = "Head-Turn Preference Paradigm",
      "singlescreen" = "Single Screen")
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
   scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

## Habituation by Caregiver Seat
```{r}
ggplot(data %>% dplyr::filter(!is.na(caregiver_seat)), 
       aes(x = testing_trial, y = looking_time, color = caregiver_seat)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Sitting Arangement",
    labels = c(
      "caregiver_lap_only" = "Caregiver Lap",
      "seat_alone_only" = "Sitting Alone",
      "changed_throughout" = "Changed Throughout")
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  theme_minimal()
```