---
title: >
  ManyBabies Habituation Project
subtitle: >
  This project aims to explore infant habituation through looking times   using a large harmonised dataset from ManyBabies 1, ManyBabies 3, and ManyBabies 4. Study-relevant materials can be found on the Open Science Framework: https://osf.io/8xm4r
author:
  
- name: Jasmin Dönicke
  affiliation: 
  - University Medical Center Mainz
  email: jasminabuaishah@gmail.com
  orcid: 0009-0006-0721-0790
      
date: 2026-03-21
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
    mainfont: "Atkinson Hyperlegible"
    # mainfont: "IBM Plex Sans"
    link-external-newwindow: true
    # code-link: true ## does seem to interfere with the title... pity.
    # code-tools: true
    theme: 
    - cosmo
    - ummz_theme.scss
execute: 
  cache: false
  echo: true
  eval: true
  output: true
  message: false
  warning: false
editor_options: 
  chunk_output_type: console
  execute:
  error: true
---

# Shape of habituation curve
Explore which functional forms best capture habituation trajectories.


##  Library
```{r}
#| results: hide
# libraries
library(dplyr)
library(ggplot2)
library(lme4)
library(splines)
library(broom.mixed)
library(flextable)
```

## Prepare Data

```{r}

#| results: hide
# set environment

# read CSV
data <- read.csv("final_merged_manybabies.csv")

# assign factors
data <- data %>%
  mutate(across(where(is.character), as.factor),
    age = cut(
      age_days,
      breaks = c(0, 90, 180, 270, 365, 455, 545, 635, 653),  # boundaries in days
      labels = c(
        "< 3 months",
        "3–6 months",
        "6–9 months",
        "9–12 months",
        "12–15 months",
        "15–18 months",
        "18–21 months",
        "> 21 months"),
      right = FALSE),
    age = dplyr::coalesce(age_group_new, age),
    age = factor(age, levels = c(
      "< 3 months",
      "3–6 months",
      "6–9 months",
      "9–12 months",
      "12–15 months",
      "15–18 months",
      "18–21 months",
      "> 21 months")),
    testing_trial = ifelse(trial_number > 0, trial_number, NA), # create variable with only testing trials (Training trials [-2, -1, 0] are excluded because they serve familiarization purposes and are not relevant for the main outcome) 
    log_looking_time = log1p(looking_time), # given the zero inflacted looking times, log1p allows them to be included in the model as log(0) is impossible
    testing_trial_sc = scale(testing_trial), # scale testing_trial to help model estimation
    age_days_sc = scale(age_days)) # scale age_days to help model estimation

# remove NA for complete-case analysis
data_model <- data %>%
  filter(!is.na(looking_time))

```

# Data validation

### Missingness on dependent variable
```{r validation}

# Distribution of zero looking times and NAs
table(is.na(data$looking_time))
table(data$looking_time == 0)

# Are zeros/NAs concentrated in later trials?
aggregate(looking_time == 0 ~ trial_number, data = data, FUN = mean)
aggregate(is.na(looking_time) ~ trial_number, data = data, FUN = mean)

```

## Simplify Data
We have roughly 55,000 observations. When testing new code, I can use this smaller sample so things run faster (= use data_sample). Once your code is ready, I will apply it to the full dataset (= use data).

```{r}
set.seed(123)

data_sample <- data %>%
  dplyr::sample_n(5000)
```

## Plot Overall Habituation Curve
The plot shows a steel early habituation (steep drop), which flattens at circa trial No. 4. This looks like a possible non-linear relationship!

```{r}
ggplot(data, aes(x = testing_trial, y = looking_time)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  theme_minimal()
```

## Plot Individual Habituation Curves

The individual plots show substantial variability in the habituation curves across individuals and time

```{r plot looking times, echo=FALSE}

set.seed(1)
ids <- sample(unique(data$unique_subject_identifier), 200) # to be able to observe trajectories for a subset of participants

df_sub <- subset(data, unique_subject_identifier %in% ids)
df_sub <- df_sub[order(df_sub$unique_subject_identifier, df_sub$testing_trial), ]

ggplot(df_sub,
       aes(x = testing_trial,
           y = looking_time,
           group = unique_subject_identifier,
           color = unique_subject_identifier)) +
  geom_line(alpha = 0.6) +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(c(0, 25))
```

## Missingness

```{r missingness, echo=FALSE}

cat_vars <- c("participant_sex", "race", "primarycaregiver_gender", "language_group", "primary_language", "caregiver_seat")
cont_vars <- c("age_days", "looking_time", "household_size", "primary_language_exposure")

cat_summary <- map_dfr(cat_vars, function(var) {
  
  x <- data[[var]]
  
  tab <- table(x, useNA = "no")  # excludes NA, added later
  
  tibble(Variable = var,
    Level = names(tab),
    N = as.integer(tab),
    Percent = 100 * N / sum(N)) %>%
    bind_rows(
      tibble(
        Variable = var,
        Level = "Missing",
        N = sum(is.na(x)),
        Percent = NA_real_))}) %>%
  flextable()

cont_summary <- map_dfr(cont_vars, function(var) {
  
  tibble(
    Variable = var,
    Mean = mean(data[[var]], na.rm = TRUE),
    SD = sd(data[[var]], na.rm = TRUE),
    Missing = sum(is.na(data[[var]])))}) %>%
  flextable()

```

## Create Overall Habituation Curve Models
To explore which functional forms best capture habituation trajectories, I model looking time as a function of trial number using a series of regression models. Specifically, I compare linear, quadratic, cubic and other shapes (using splines) relationships the variables.

First, I estimate a linear model to assess whether looking time decreases across trials. I then extend this model by including a quadratic term to capture potential curvature in the habituation trajectory. In addition, I fit flexible non-linear models using spline functions to allow for more complex shapes. Importantly, I always use mixed-models with infant-level random intercepts, to account for repeated measurements within infants.

```{r}
model_linear_mixed <- lmer(looking_time ~ testing_trial + (1 | unique_subject_identifier), data = data_model)

model_quadratic_mixed <- lmer(
  looking_time ~ testing_trial + I(testing_trial^2) + (1 | unique_subject_identifier),
  data = data_model)

model_spline_mixed <- lmer(
  looking_time ~ ns(testing_trial, df = 3) + (1 | unique_subject_identifier),
  data = data_model)
```

### Linear Model
- Looking-time decreases significantly across trials (Slope = -0.27 [95% CI -0.28,    -0.27], t = -72.79), which tells us: habituation is happening. 
- On average, looking-time decreased by approximately 0.27 seconds per trial. 
- There is substantial variability in baseline looking- times across infants, supporting the inclusion of random intercepts.
- The average looking-time in the first trial was 9.2 seconds.
```{r}
summary(model_linear_mixed)
broom.mixed::tidy(model_linear_mixed, conf.int = TRUE)
```

### Quadratic Model
- There is a significant decrease in looking time across trials (Slope = -0.49 [95% CI -0.516, -0.459], t = -33.69) and a significant positive quadratic effect (b = 0.013, t = 15.33). 
- This indicates that looking-time decreased rapidly at early trials and then leveled off over time, consistent with a typical habituation trajectory. 
```{r}
summary(model_quadratic_mixed)
broom.mixed::tidy(model_quadratic_mixed, conf.int = TRUE)
```

### Spline Model
- There is a significant nonlinear decrease in looking time across trials. 
- Splines cannot be interpreted as easily.
- The spline model provided the best fit to the data compared to linear and quadratic models, indicating that habituation follows a nonlinear trajectory characterized by a rapid initial decrease and subsequent leveling off. This pattern is consistent with typical habituation processes.
```{r}
# A mixed-effects model using natural splines revealed. 
summary(model_spline_mixed)
broom.mixed::tidy(model_spline_mixed, conf.int = TRUE)
```

```{r}
# Plot Spline Model
ggplot(data_model, aes(x = testing_trial, y = looking_time)) +
  geom_smooth(method = "lm", 
              formula = y ~ splines::ns(x, df = 3), 
              se = TRUE) +
  theme_minimal()
```

```{r}
# Plot Spline Model on top of observed data
ggplot(data, aes(x = testing_trial, y = looking_time)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  geom_smooth(method = "lm", 
              formula = y ~ splines::ns(x, df = 3), 
              se = TRUE) +
  theme_minimal()
# Here, I am showing the average observed habituation pattern and overlaying a smooth model-based curve to see how well a nonlinear trajectory captures it. 
```


## Compare  Overall Habituation Curve Model Fit 

Here, I compare models based on goodness-of-fit criteria (e.g. AIC).

```{r}
AIC(model_linear_mixed, model_quadratic_mixed, model_spline_mixed)
BIC(model_linear_mixed, model_quadratic_mixed, model_spline_mixed)
```

Here, I compare models by visual inspection of fitted trajectories. This allows me to evaluate whether habituation follows a linear decline, shows rapid early decreases with later stabilization, or exhibits more complex patterns.
```{r}
data_summary <- data_model %>%
  group_by(testing_trial) %>%
  summarise(
    mean_lt = mean(looking_time, na.rm = TRUE),
    sd_lt = sd(looking_time, na.rm = TRUE),
    n = sum(!is.na(looking_time)),
    se_lt = sd_lt / sqrt(n),
    ci_low = mean_lt - 1.96 * se_lt,
    ci_high = mean_lt + 1.96 * se_lt,
    .groups = "drop"
  )

# Linear Model
ggplot(data_summary, aes(x = testing_trial, y = mean_lt)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, level = 0.95) +
  labs(
    x = "Trial Number",
    y = "Mean Looking Time",
    title = "Linear Habituation Trajectory"
  ) +
  coord_cartesian(ylim = c(5, 10)) +
   scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()

# Quadratic Model
ggplot(data_summary, aes(x = testing_trial, y = mean_lt)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2), se = TRUE, level = 0.95) +
  labs(
    x = "Trial Number",
    y = "Mean Looking Time",
    title = "Quadratic Habituation Trajectory"
  ) +
  coord_cartesian(ylim = c(5, 10)) +
   scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()

# Spline Model
ggplot(data_summary, aes(x = testing_trial, y = mean_lt)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, df = 3), se = TRUE, level = 0.95) +
  labs(
    x = "Trial Number",
    y = "Mean Looking Time",
    title = "Spline Model Habituation Trajectory"
  ) +
  coord_cartesian(ylim = c(5, 10)) +
   scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

All models in one plot
```{r}
ggplot(data_summary, aes(x = testing_trial, y = mean_lt)) +
  geom_point(size = 2.2, color = "black") +
  
  geom_smooth(
    aes(color = "Linear", linetype = "Linear"),
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    linewidth = 1
  ) +
  geom_smooth(
    aes(color = "Quadratic", linetype = "Quadratic"),
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE,
    linewidth = 1
  ) +
  geom_smooth(
    aes(color = "Spline", linetype = "Spline"),
    method = "lm",
    formula = y ~ splines::ns(x, df = 3),
    se = FALSE,
    linewidth = 1
  ) +

  scale_color_manual(
    name = "Model",
    values = c(
      "Linear" = "darkgreen",
      "Quadratic" = "blue",
      "Spline" = "red"
    )
  ) +
  scale_linetype_manual(
    name = "Model",
    values = c(
      "Linear" = "dashed",
      "Quadratic" = "solid",
      "Spline" = "dotted"
    )
  ) +
  scale_x_continuous(breaks = 1:16) +
  labs(
    x = "Trial Number",
    y = "Mean Looking Time",
    title = "Habituation Trajectories by Model"
  ) +
  coord_cartesian(ylim = c(5, 10)) +
  theme_minimal()
```


# Individual differences
Identify factors that explain variability across infants.

- 2.1 What factors moderate differences?
- 2.2 Which moderators are most important?
- 2.3 Do moderators relate to different aspects?
- 2.4 Are there identifiable subgroups?

## Habituation by Age
I will use the age variable that I previously created by harmonizing the two variables age_group_new and age_days
```{r}
ggplot(data %>% dplyr::filter(testing_trial > 0), 
       aes(x = testing_trial, y = looking_time, color = age)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Age"
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```


The plot looks strange: Lets check, how many observations are in each age group in the testing trials!
```{r}
data %>%
  dplyr::filter(testing_trial > 0) %>%
  dplyr::count(age)
```

There are few observations in 15-18 and 18-21 months. I will remove NA category for plotting, and add 15-21 months into a single category to make the sample bigger.
```{r}
data <- data %>%
  mutate(new_age = recode(age,
                              "< 3 months" = "< 3 months",
                              "3–6 months" = "3–6 months",
                              "6–9 months" = "6–9 months",
                              "9–12 months" = "9–12 months",
                              "12–15 months" = "12–15 months",
                              "15–18 months" = "15–21 months",
                              "18–21 months" = "15–21 months",
                              "21+ months" = "21+ months"))

# check new age-variable
data %>%
  dplyr::filter(testing_trial > 0) %>%
  dplyr::count(new_age)
```

Plot again with new variable and without NA group (16 observations), with 95% confidence intervals 
```{r}
ggplot(data %>% dplyr::filter(testing_trial > 0, !is.na(age) ), 
       aes(x = testing_trial, y = looking_time, color = new_age)) +
  geom_smooth(se = TRUE, level = 0.95) +
  scale_color_discrete(
    name = "Age"
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

## Habituation by Sex of Participant
There seems to be no sex difference overall! Maybe if we stratify the data by other factors later on.
```{r}
ggplot(data %>% dplyr::filter(testing_trial > 0, !is.na(participant_sex) ), 
       aes(x = testing_trial, y = looking_time, color = participant_sex)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Sex"
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

## Habituation by Gender of Primary Caregiver
```{r}
# check amount of data per category
data %>%
dplyr::filter(testing_trial > 0) %>%
dplyr::count(primarycaregiver_gender)
# "other" and "OTHER" are just 75 observations, compared to 44000 F and 4500 M, remove "other" due to severely missing data for plotting

ggplot(data %>% dplyr::filter(
         !is.na(primarycaregiver_gender),   
         !primarycaregiver_gender %in% c("other", "OTHER")), 
       aes(x = testing_trial, y = looking_time, color = primarycaregiver_gender)) +
  geom_smooth(se = TRUE, level = 0.95) +
  scale_color_discrete(
    name = "Primary Caregiver Gender"
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

## Habituation by Language Group
```{r}
# check amount of data per category
data %>%
dplyr::filter(testing_trial > 0) %>%
dplyr::count(language_group)
# "other" and "OTHER" are just 71 observations, remove them for plotting

ggplot(data %>% dplyr::filter(!is.na(language_group),
                              !language_group %in% c("other")), 
       aes(x = testing_trial, y = looking_time, color = language_group)) +
          geom_smooth(se = TRUE, level = 0.95) +
          scale_color_discrete(
            name = "Lanuage Group"
          ) +
          labs(
            x = "Trial Number",
            y = "Looking Time (in seconds)",
            title = "Looking Time Across Trials"
          ) +
          scale_x_continuous(
            breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
          ) +
          theme_minimal()
```



# Methodological effects
Assess how methodological choices influence results.

- 3.1 Do differences contribute to variability?
- 3.2 Do they affect interpretation?
- 3.3 Do they affect reliability?


## Habituation by Trial Type
```{r}
# check amount of data per category
data %>%
dplyr::filter(testing_trial > 0) %>%
dplyr::count(trial_type)


ggplot(data, aes(x = testing_trial, y = looking_time, color = trial_type)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Trial Type",
    labels = c(
      "ADS" = "Adult speech",
      "IDS" = "Infant speech",
      "consistent" = "Consistent",
      "inconsistent" = "Inconsistent",
      "nonsocial" = "Non-Social",
      "social" = "Social"
    )
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

## Habituation by Project ID
```{r}
ggplot(data, aes(x = testing_trial, y = looking_time, color = project_id)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Project ID") +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

#### Habituation by Method
```{r}
ggplot(data, aes(x = testing_trial, y = looking_time, color = method)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Method",
    labels = c(
      "eyetracking" = "Eyetracking",
      "hpp" = "Head-Turn Preference Paradigm",
      "singlescreen" = "Single Screen")
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
   scale_x_continuous(
    breaks = seq(1, max(data$testing_trial, na.rm = TRUE), by = 1)
  ) +
  theme_minimal()
```

## Habituation by Caregiver Seat
```{r}
ggplot(data %>% dplyr::filter(!is.na(caregiver_seat)), 
       aes(x = testing_trial, y = looking_time, color = caregiver_seat)) +
  geom_smooth(se = TRUE) +
  scale_color_discrete(
    name = "Sitting Arangement",
    labels = c(
      "caregiver_lap_only" = "Caregiver Lap",
      "seat_alone_only" = "Sitting Alone",
      "changed_throughout" = "Changed Throughout")
  ) +
  labs(
    x = "Trial Number",
    y = "Looking Time (in seconds)",
    title = "Looking Time Across Trials"
  ) +
  theme_minimal()
```